What is happening on Twitter? A framework for student research projects with tweets

### 2020-09-11 Rmd/jse-special-issue-2020-october.Rmd
replaced "latent Dirichlet allocation" with "LDA" abbreviation

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, 
we analyzed tweets before, during, and 
after Memorial Day (May 25, 2020). 
We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time 
on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of 
tweets.

@@

Readers may need to slightly amend the above line to conform to requirements of their operating system's `crontab`. Readers who use Mac OS may proceed as we did, while those with Windows operating systems may consider using the R package "taskscheduleR" to schedule API queries via the Windows task scheduler [@taskscheduleR].

@@

The next task is to wrangle the tweet JSON into a data structure suitable for latent Dirichlet allocation modeling. 
We used functions from the `rtweet` R package to parse tweet JSON into a data frame. 
We then divided tweet text into words with functions from the `tidytext` R package. We 
discarded commonly used "stop words" and emojis.

Latent Dirichlet allocation model fitting requires that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

## Latent Dirichlet allocation

Latent Dirichlet allocation is a statistical method for inferring latent (unobservable) topics (or themes)
from a large corpus (or collection) of documents [@blei2003latent].
We pretend that there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may 
refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic.

@@

The goal for latent Dirichlet allocation is to infer both the distribution over topics and 
the topics [@blei2003latent]. A topic, in this setting, is a distribution over 
the vocabulary (the collection of all words in a corpus).

@@

Latent Dirichlet allocation modeling and related methods are a major research area in the 
quantitative social sciences. Advanced students with interest in statistical computing might
compare inferential methods for topic models. Those with interests in event detection
and time series analysis could build on the findings of our student by explicitly accounting
for topic evolution with dynamic topic models [@blei2006dynamic].

>>>>

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, 
we analyzed tweets before, during, and 
after Memorial Day (May 25, 2020). 
We fitted LDA models for each of three
distinct five-minute periods. The first period began at noon Eastern time 
on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of 
tweets.

@@

Readers may need to slightly amend the above line to conform to requirements of 
their operating system's software. Readers who use Mac OS may proceed as we did, 
while those with Windows operating systems may consider using the R package 
"taskscheduleR" to schedule API queries via the Windows task scheduler [@taskscheduleR].

@@

The next task is to wrangle the tweet JSON into a data structure suitable for LDA modeling. 
We used functions from the `rtweet` R package to parse tweet JSON into a data frame. 
We then divided tweet text into words with functions from the `tidytext` R package. We 
discarded commonly used "stop words" and emojis.

LDA model fitting requires that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

## Latent Dirichlet Allocation

LDA is a statistical method for inferring latent (unobservable) topics (or themes)
from a large corpus (or collection) of documents [@blei2003latent].
We pretend that there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may 
refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic.

@@

The goal for LDA is to infer both the distribution over topics and 
the topics [@blei2003latent]. A topic, in this setting, is a distribution over 
the vocabulary (the collection of all words in a corpus).

@@

LDA modeling and related methods are a major research area in the 
quantitative social sciences. Advanced students with interest in statistical computing might
compare inferential methods for topic models. Those with interests in event detection
and time series analysis could build on the findings of our student by explicitly accounting
for topic evolution with dynamic topic models [@blei2006dynamic].
